Realizability-Preserving DG-IMEX Method for the Two-Moment Model of Fermion Transport \tnoteref{support}

### 2018-06-22 Documents/M1/realizableFermionicM1/sections/appendix.tex
Reworked Appendix B

For the analysis of positivity-preserving IMEX schemes, additional coefficients are defined \cite{hu_etal_2018} (cf. Eq.~\eqref{eq:imexStagesRewrite}).  
First, let
\begin{equation}
  b_{ii} = \f{1}{a_{ii}}, \quad
  b_{ij} = -\f{1}{a_{ii}}\sum_{l=j}^{i-1}a_{il}b_{lj}, \quad
  \tilde{b}_{ij} = -\f{1}{a_{ii}}\Big(\tilde{a}_{ij}+\sum_{l=j+1}^{i-1}a_{il}\tilde{b}_{lj}\Big).  
\end{equation}
Then, for IMEX schemes of Type~A \cite{dimarcoPareschi2013},
\begin{equation}
  \begin{aligned}
    c_{i0} &= 1-\sum_{j=1}^{i-1}\sum_{l=j}^{i-1}a_{il}b_{lj}, \quad &
    c_{ij} &= \sum_{l=j}^{i-1}a_{il}b_{lj}, \\
    \tilde{c}_{i0} &= 0, \quad &
    \tilde{c}_{ij} &= \tilde{a}_{ij} + \sum_{l=j+1}^{i-1}a_{il}\tilde{b}_{lj}.
  \end{aligned}
\end{equation}
while for IMEX schemes of Type~ARS \cite{ascher_etal_1997}
\begin{equation}
\begin{aligned}
  c_{i0} &= 1-\sum_{j=2}^{i-1}\sum_{l=j}^{i-1}a_{il}b_{lj}, \quad &
  c_{ij} &= \sum_{l=j}^{i-1}a_{il}b_{lj} \\
  \tilde{c}_{i0} &= \tilde{a}_{i1}+\sum_{j=2}^{i-1}a_{ij}\tilde{b}_{j1}, \quad &
  \tilde{c}_{ij} &= \tilde{a}_{ij}+\sum_{l=j+1}^{i-1}a_{il}\tilde{b}_{lj}.  
\end{aligned}
\end{equation}

@@

\newpage 
\section{Construction of PD-ARS}
\label{app:PD-ARS}

It can be shown that no 2-stage positivity-preserving IMEX exist \cite{hu_etal_2018}.
Hence we considered 3-stage IMEX scheme.
There are two kinds of implicit DIRK scheme, Type A and Type ARS.
Here we discuss the simpler case, Type ARS, first.
The Butcher tableau for ARS type IMEX scheme is
\begin{equation}
  \begin{array}{c | c c c}
  	         0           & 0                    & 0                    & 0                    \\
  	\tilde{c}_{2} & \tilde{a}_{21}       & 0                    & 0                    \\
  	\tilde{c}_{3} & \tilde{a}_{31}       & \tilde{a}_{32}       & 0                    \\ \hline
  	                     & \tilde{w}_{1} & \tilde{w}_{2} & \tilde{w}_{3}
  \end{array}
  \qquad
  \begin{array}{c | c c c}
  	     0       & 0            & 0            & 0            \\
  	c_{2} & 0            & a_{22}       & 0            \\
  	c_{3} & 0            & a_{32}       & a_{33}       \\ \hline
  	             & w_{1} & w_{2} & w_{3}
  \end{array}.
\end{equation}
The properties a desired IMEX scheme should have are: 2nd-order accurate in the streaming, globally stiffly accurate, positivity-preserving, diffusion accurate and an affordable time-step restriction.
Then the problem that find such a scheme can be rewritten as an optimization problem:\\
Find $(\tilde{a}_{21}, \tilde{a}_{31}, \tilde{a}_{32}, \tilde{w}_{1}, \tilde{w}_{2}, \tilde{w}_{3}, a_{22}, a_{32}, a_{33}, w_{1}, w_{2}, w_{3})$ such that
\begin{align*}
\text{maximize} ~ c_{sch} = &\min \{ \dfrac{c_{20}}{\tilde{c}_{20}}, \dfrac{c_{30}}{\tilde{c}_{30}}, \dfrac{c_{32}}{\tilde{c}_{32}}\}\\
 \text{under these constraints:} \qquad & \\
 \text{(GSA)} \qquad\ &\tilde{w}_{i} = \tilde{a}_{3i}, \quad w_{i} = a_{3i}, ~\text{for} ~ i = 1,2,3\\
 \text{(consistency)} \qquad\ & \sum_{i=1}^{3}  \tilde{w}_{i} = 1, \quad \sum_{i=1}^{3} w_{i} = 1\\ 
 \text{(2nd-order streaming accuracy)} \qquad\ & \sum_{i=1}^{3}  \tilde{w}_{i}\tilde{c}_{i} = 1/2, \\
 \text{(positivity-preserving)} \qquad & a_{ii} > 0, c_{i0} \geq 0, \tilde{c}_{i0} \geq 0, ~\text{for} ~ i = 2,3 \\
 & c_{32} \geq 0, \tilde{c}_{32} \geq 0, \\
 \text{(diffusion accurate)} \qquad & e^{T}_iA^{-1}\tilde{A}e = 1, ~\text{for} ~ i = 2,3, \\
 \text{(affordable)} \qquad & c_{sch} > 0.
\end{align*}
Imposing those equality constraints (including all the constraints exclude those from the positivity-preserving and affordable requirements), the Butcher tableau can be simplified as:
\begin{equation}
  \begin{array}{c | c c c}
  	     0       & 0            & 0 & 0 \\
  	\frac{1}{2x} & \frac{1}{2x} & 0 & 0 \\
  	     1       & 1-x          & x & 0 \\ \hline
  	             & 1-x          & x & 0
  \end{array}
  \qquad
  \begin{array}{c | c c c}
  	     0       & 0 & 0            & 0 \\
  	\frac{1}{2x} & 0 & \frac{1}{2x} & 0 \\
  	     1       & 0 & 1-y          & y \\ \hline
  	             & 0 & 1-y          & y
  \end{array}.
\end{equation}
And the rest inequalities and the problem were left to be: \\
Find $(x,y)$ such that
\begin{align*}
\text{maximize} ~ c_{sch} = &\min \{ 2x, \dfrac{1-2x(1-y)}{y-x}, 2(1-y)\}\\
 \text{under these constraints:} \qquad & x > 0, ~ 1 \geq y > 0, \\
 & x \leq y, ~\dfrac{1}{2x} \geq 1-y.
\end{align*}
A solution family is $x=\frac{1}{2}$, $y=\frac{1}{2} + \epsilon$ with $\epsilon \in [0, \frac{1}{2})$ and $c_{sch} = 1 - 2\epsilon$.
The PD-ARS scheme follows.
It's clear that $\epsilon = 0$ and $c_{sch} = 1$ is the most optimized solution.

>>>>

For the analysis of positivity-preserving IMEX schemes, additional coefficients are defined \cite{hu_etal_2018} (cf. Eq.~\eqref{eq:imexStagesRewrite}).  
First, let
\begin{equation}
  b_{ii} = \f{1}{a_{ii}}, \quad
  b_{ij} = -\f{1}{a_{ii}}\sum_{l=j}^{i-1}a_{il}b_{lj}, \quad
  \tilde{b}_{ij} = -\f{1}{a_{ii}}\Big(\tilde{a}_{ij}+\sum_{l=j+1}^{i-1}a_{il}\tilde{b}_{lj}\Big).  
\end{equation}
Then, for IMEX schemes of Type~A \cite{dimarcoPareschi2013},
\begin{equation}
  \begin{aligned}
    c_{i0} &= 1-\sum_{j=1}^{i-1}\sum_{l=j}^{i-1}a_{il}b_{lj}, \quad &
    c_{ij} &= \sum_{l=j}^{i-1}a_{il}b_{lj}, \\
    \tilde{c}_{i0} &= 0, \quad &
    \tilde{c}_{ij} &= \tilde{a}_{ij} + \sum_{l=j+1}^{i-1}a_{il}\tilde{b}_{lj}.
  \end{aligned}
  \label{eq:positivityCoefficientsA}
\end{equation}
while for IMEX schemes of Type~ARS \cite{ascher_etal_1997}
\begin{equation}
  \begin{aligned}
    c_{i0} &= 1-\sum_{j=2}^{i-1}\sum_{l=j}^{i-1}a_{il}b_{lj}, \quad &
    c_{ij} &= \sum_{l=j}^{i-1}a_{il}b_{lj} \\
    \tilde{c}_{i0} &= \tilde{a}_{i1}+\sum_{j=2}^{i-1}a_{ij}\tilde{b}_{j1}, \quad &
    \tilde{c}_{ij} &= \tilde{a}_{ij}+\sum_{l=j+1}^{i-1}a_{il}\tilde{b}_{lj}.  
  \end{aligned}
  \label{eq:positivityCoefficientsARS}
\end{equation}

@@

\newpage 
\section{Construction of IMEX Scheme PD-ARS}
\label{app:PD-ARS}

Here we construct a three-stage PD-IMEX scheme of Type~ARS, conforming to Definition~\ref{def:PD-IMEX}.  
We refer to the resulting IMEX scheme as PD-ARS.  
The double Butcher tableau is
\begin{equation}
  \begin{array}{c | c c c}
  	         0           & 0                 & 0                   & 0                    \\
  	\tilde{c}_{2} & \tilde{a}_{21} & 0                   & 0                    \\
  	\tilde{c}_{3} & \tilde{a}_{31} & \tilde{a}_{32} & 0                    \\ \hline
  	                   & \tilde{a}_{31} & \tilde{a}_{32} & 0
  \end{array}
  \qquad
  \begin{array}{c | c c c}
  	     0  & 0  & 0         & 0            \\
  	c_{2} & 0 & a_{22} & 0            \\
  	c_{3} & 0 & a_{32} & a_{33}       \\ \hline
  	         & 0 & a_{32} & a_{33}
  \end{array}
\end{equation}
The problem is then to find the coefficients $\{ \tilde{a}_{21}, \tilde{a}_{31}, \tilde{a}_{32}, a_{22}, a_{32}, a_{33} \}$ satisfying the constraints in Definition~\ref{def:PD-IMEX} while maximizing
\begin{equation}
  c_{\mbox{\tiny Sch}} = \min \Big\{\, \dfrac{c_{20}}{\tilde{c}_{20}},\, \dfrac{c_{30}}{\tilde{c}_{30}},\, \dfrac{c_{32}}{\tilde{c}_{32}} \,\Big\}.  
\end{equation}
%\begin{align*}
%\text{maximize} ~ c_{sch} = &\min \{ \dfrac{c_{20}}{\tilde{c}_{20}}, \dfrac{c_{30}}{\tilde{c}_{30}}, \dfrac{c_{32}}{\tilde{c}_{32}}\}\\
% \text{under these constraints:} \qquad & \\
% \text{(GSA)} \qquad\ &\tilde{w}_{i} = \tilde{a}_{3i}, \quad w_{i} = a_{3i}, ~\text{for} ~ i = 1,2,3\\
% \text{(consistency)} \qquad\ & \sum_{i=1}^{3}  \tilde{w}_{i} = 1, \quad \sum_{i=1}^{3} w_{i} = 1\\ 
% \text{(2nd-order streaming accuracy)} \qquad\ & \sum_{i=1}^{3}  \tilde{w}_{i}\tilde{c}_{i} = 1/2, \\
% \text{(positivity-preserving)} \qquad & a_{ii} > 0, c_{i0} \geq 0, \tilde{c}_{i0} \geq 0, ~\text{for} ~ i = 2,3 \\
% & c_{32} \geq 0, \tilde{c}_{32} \geq 0, \\
% \text{(diffusion accurate)} \qquad & e^{T}_iA^{-1}\tilde{A}e = 1, ~\text{for} ~ i = 2,3, \\
% \text{(affordable)} \qquad & c_{sch} > 0.
%\end{align*}
Imposing the equality constraints (i.e., Eqs.~\eqref{eq:implicitConsistency}, \eqref{eq:orderConditionsEx}, and \eqref{eq:diffusionCondition}), the double Butcher tableau can be written in terms of two parameters ($x$ and $y$) as
\begin{equation}
  \begin{array}{c | c c c}
  	     0       & 0            & 0 & 0 \\
  	\frac{1}{2x} & \frac{1}{2x} & 0 & 0 \\
  	     1       & 1-x          & x & 0 \\ \hline
  	             & 1-x          & x & 0
  \end{array}
  \qquad
  \begin{array}{c | c c c}
  	     0       & 0 & 0            & 0 \\
  	\frac{1}{2x} & 0 & \frac{1}{2x} & 0 \\
  	     1       & 0 & 1-y          & y \\ \hline
  	             & 0 & 1-y          & y
  \end{array}
\end{equation}
Computing the relevant coefficients in Eq.~\eqref{eq:positivityCoefficientsARS}, we find $c_{20}=1$, $c_{30}=1-2x(1-y)$, $c_{32}=2x(1-y)$, $\tilde{c}_{20}=\f{1}{2x}$, $\tilde{c}_{30}=(y-x)$, and $\tilde{c}_{32}=(1-y)$, so that
\begin{equation}
  c_{\mbox{\tiny Sch}} = \min \Big\{\, 2x,\, \f{1-2x(1-y)}{y-x},\, 2(1-y) \,\Big\}.  
\end{equation}
The positivity-preserving property follows from imposing the inequality constraints $a_{22},a_{33}>0$, $c_{20},c_{30},c_{32}\ge0$, and $\tilde{c}_{20},\tilde{c}_{30},\tilde{c}_{32}\ge0$, which imply
\begin{equation}
  0 < x \le y
  \quad\text{and}\quad
  0 < y \le 1.
\end{equation}
We chose $x=\f{1}{2}$, so that the explicit part of the IMEX scheme is equivalent to the optimal second-order SSP-RK scheme in \cite{gottlieb_etal_2001} (SSPRK2 in \ref{app:butcherTables}).  
Then, $y=\f{1}{2} + \epsilon$, where $\epsilon \in [0, \frac{1}{2})$ and $c_{\mbox{\tiny Sch}} = 1 - 2\epsilon$, results in the PD-ARS IMEX scheme.  
Setting $\epsilon = 0$ gives the optimal scheme with $c_{\mbox{\tiny Sch}} = 1$.

%It remains to find And the rest inequalities and the problem were left to be: \\
%Find $(x,y)$ such that
%\begin{align*}
%\text{maximize} ~ c_{sch} = &\min \{ 2x, \dfrac{1-2x(1-y)}{y-x}, 2(1-y)\}\\
% \text{under these constraints:} \qquad & x > 0, ~ 1 \geq y > 0, \\
% & x \leq y, ~\dfrac{1}{2x} \geq 1-y.
%\end{align*}
%A solution family is $x=\frac{1}{2}$, $y=\frac{1}{2} + \epsilon$ with $\epsilon \in [0, \frac{1}{2})$ and $c_{sch} = 1 - 2\epsilon$.
%The PD-ARS scheme follows.
%It's clear that $\epsilon = 0$ and $c_{sch} = 1$ is the most optimized solution.
